---
title: "Case Study 1"
author: "Crystal Todd"
date: "October 31, 2016"
output:
  html_document:
    keep_md: yes
---

## Introduction

In regards to the GDP of several countries, we would like to analyze the GDP of countries against their income groups. This analysis will help define the economic health of countries compared to other demographic data for the countries.

## Data Downloading, Tidying, and Merging
The process used to clean the data began with downloading the data provided, unsure of the format or condition of the data. First, we must load the packages required for this analysis and download the two data sets from the given URLs.
```{r}
# Load Needed Packages
library(repmis)
library(dplyr)
library(ggplot2)

# Set the Working Directory
setwd("C:/Users/cltod/Documents/MSDS 6306 Doing Data Science/Unit 8/")

# Data Set 1
URL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv"
download.file(URL, destfile = "./getdataGDP.csv")
rm(URL)
GDP <- read.csv("getdataGDP.csv", header = FALSE, stringsAsFactors = FALSE) # Load Data Set into R from Working Directory. Use header = FALSE because we do not know the format and condition of the data set.

# Data Set 2
URL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv"
download.file(URL, destfile = "./getdataEDSTATS.csv")
rm(URL)
EDSTATS <- read.csv("getdataEDSTATS.csv", header = FALSE, stringsAsFactors = FALSE) # Load Data Set into R from Working Directory. Use header = FALSE because we do not know the format and condition of the data set.
```

First, we cleaned the GDP data by looking at several characteristics of the data.

```{r}
dim(GDP)
str(GDP)
head(GDP)
```
The header of the GDP data appears to have some column headers in the fifth row. We renamed the columns we could quickly identify and remove the top five rows.
```{r}
GDP <- rename(GDP, COUNTRY_CODE = V1, RANKING = V2, ECONOMY = V4, USD_MIL = V5)
str(GDP)
tail(GDP)
GDP <- GDP[6:nrow(GDP),]
head(GDP)
```
Next, there are six columns we cannot quickly identify the contents of the columns. We looked at these columns to find if they are completely empty or not.
```{r}
table(GDP$V3)
table(GDP$V6)
table(GDP$V7)
table(GDP$V8)
table(GDP$V9)
table(GDP$V10)
```

Only the column V6 contained any data with some kind of comment in the form of a through f; the other columns contained no data and were deleted from the data set with V6 being renamed to the column name COMMENTS.
```{r}
GDP <- select(GDP, COUNTRY_CODE, RANKING, ECONOMY, USD_MIL, COMMENTS = V6)
```

Previously, we looked at the tail of the data set, and it appeared that there were at least six rows at the end of the data that had no data in them.

```{r}
table(GDP$COUNTRY_CODE) #There are 98 blank values aside from the first five rows deleted from the beginning of analysis.

nulls <- GDP[GDP$COUNTRY_CODE == "",] # Splitting null from non-null Country Code values to examine.
nonull <- GDP[GDP$COUNTRY_CODE != "",]

table(nulls$RANKING)

## DATA HIDING IN RANKING COLUMN
# .. Not available.
# a. Includes Former Spanish Sahara.
# b. Excludes South Sudan
# c. Covers mainland Tanzania only.
# d. Data are for the area controlled by the government of the Republic of Cyprus.
# e. Excludes Abkhazia and South Ossetia.
# f. Excludes Transnistria.
# Note: Rankings include only those economies with confirmed GDP estimates. Figures in italics are for 2011 or 2010.

table(nulls$ECONOMY)
table(nulls$USD_MIL)
table(nulls$V6)
```

Since those 98 blank rows have been examined and the important information regarding a through f has been identified, we can delete those rows in the data set.

```{r}
# For all of the 98 blank values for COUNTRY_CODE, the other columns are blank. Deleting those 98 rows from the data set GDP.
GDP <- GDP[GDP$COUNTRY_CODE != "",]
rm(nulls, nonull)
```

From the information in the RANKING column, we know that values that have ".." in the USD_MIL column are considered not applicable. This assumption was applied to the data.

```{r}
for(j in 1:nrow(GDP))
{
  if(GDP$USD_MIL[j] == "..")
    GDP$USD_MIL[j] = NA
}
```

In the data, there were commas separating the thousands place in USD_MIL, and the RANKING column is considered a character field. These must be changed to numerics and any commas removed.

```{r}
GDP$RANKING <- as.numeric(gsub("[^[:digit:]]","", GDP$RANKING))
GDP$USD_MIL <- as.numeric(gsub("[^[:digit:]]","", GDP$USD_MIL))
```

There are values in the RANKING and USD_MIL columns that did not have data values or were NA. These values must be removed to allow for correct analysis.

```{r}
nrow(GDP[is.na(GDP$RANKING),])
# There are 122 NA records in the RANKING column.
nrow(GDP[is.na(GDP$USD_MIL) | GDP$USD_MIL == "",])
# There are 135 records in USD_MIL column that are either NA or do not have a value.
# Remove where there is no RANKING
GDP <- GDP[!is.na(GDP$RANKING),]
```

The GDP data is clean for the analysis portion of the project. Now, we must clean the EDSTATS data set. First, we look at the characteristics of the data set.

```{r}
dim(EDSTATS)
head(EDSTATS)
EDSTATS <- read.csv("getdataEDSTATS.csv", header = TRUE, stringsAsFactors = FALSE) #Notice that the first row of the data set has apparent headers. We can reload the data with header = TRUE.
head(EDSTATS)
str(EDSTATS)
table(EDSTATS$CountryCode)
```

There are no rows that have blank CountryCode so there is no need to look into deleting rows. However, there are many columns to this data set, and we would only like to analyze the CountryCode, Long.Name, and Income.Group fields.

```{r}
EDSTATS <- select(EDSTATS, CountryCode, Long.Name, Income.Group)
table(EDSTATS$Long.Name)
table(EDSTATS$Income.Group)
# Notice, there are 24 blanks in Income.Group column.
```

The data are ready to analyze.

## Analysis

## Part 1: Merging Data
### Merge the data based on the country shortcode. How many of the IDs match?

```{r}
merged <- merge(GDP, EDSTATS, by.x = "COUNTRY_CODE", by.y = "CountryCode", all = FALSE)
nrow(merged)
```

There are 189 Country shortcode IDs that matched between the GDP and the Education data.

## Part 2: Sorting Data
### Sort the data frame in ascending order by GDP (so United States is last). What is the 13th country in the resulting data frame?

```{r}
mergedsort <- merged[order(merged$USD_MIL, decreasing = FALSE),]
mergedsort[13,]
```

The 13th country with the lowest GDP in millions of US dollars is St. Kitts and Nevis (Country Code KNA).


## Part 3: Average GDP Rankings
### What are the average GDP rankings for the "High Income: OECD" and "High Income: OECD" groups?

```{r}
mean(mergedsort$RANKING[mergedsort$Income.Group == "High income: OECD"])
mean(mergedsort$RANKING[mergedsort$Income.Group == "High income: nonOECD"])
```

The average GDP ranking for the "High income: OECD" group is 32.96667, and the average GDP ranking for the "High income: nonOECD" group is 91.91304.

## Part 4: Plotting Data
### Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group.

```{r}
ggplot(mergedsort)+ geom_point(aes(x=ECONOMY,y=USD_MIL,colour=Income.Group))+labs(x = "Country")
```

## Part 5: Quantiling Data
### Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are Lower middle income but among the 38 nations with highest GDP?

```{r}
mergedsort$QUANTILE <- cut(mergedsort$RANKING, breaks = 5)
table(mergedsort$QUANTILE, mergedsort$Income.Group)
mergedsort[ mergedsort$Income.Group == "Lower middle income" & mergedsort$RANKING <= 38,]
```

There are five countries that are Lower middle income but among the 38 nations with highest GDP. These countries are Arab Republic of Egypt, Thailand, Indonesia, India, and China.

## Conclusion

This analysis provided useful outcomes in regards to the countries with higher GDP but have a lower middle income group. This analysis also resulted in a plot of countries by their GDP and Income Group.